---
title: "JMML: scRNAseq surface marker"
author: "by Alex"
output: 
  html_document:
  toc: TRUE
toc_float: FALSE
editor_options: 
  chunk_output_type: console
---
  
  
  <style type="text/css">
  .main-container {
    max-width: 1800px;
    margin-left: 20px;
    margin-right: auto;
  }
.toc-content {
  max-width: 1800px;
  margin-left: 50px;
  margin-right: auto;
}

div {
  
  margin-left: 20px;
}


hr.new1 {
  border-top: 1px solid #84a8e0;
}



</style>
  
  # Results {.tabset}
  
  
  <div style="background-color: #f0f5c1;">
  
  
  </div>
  
  <hr>
  
  
  
```{r setup, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, include = FALSE , cache.lazy = FALSE )

library ( dplyr)
library(scales)
library("RColorBrewer")
library("ggplot2")
library(gplots)
library(ggrepel)
library(Hmisc)
library(stringr)
library(gridExtra)
library(knitr)
library(kableExtra)
library ( openxlsx)
library ( ggbeeswarm )
library(Seurat)
library ( HGNChelper )

lapply(c("ggraph","igraph","tidyverse", "data.tree"), library, character.only = T)




options(scipen=999)
getPalette = colorRampPalette(brewer.pal(9, "Set1")) # expand color pallete
source ( "accessory.R")

```


```{r}

# after preprocessing and merging will be stored here
preprocessed_dir = "/projects/lab.mis/elliot/JMML_project1/pre_processed/"
dir.create ( preprocessed_dir )
filename = paste0(preprocessed_dir, "HTAN.rds" )

normalizeded_dir = "/projects/lab.mis/elliot/JMML_project1/normalized/"
dir.create ( normalizeded_dir )
filename_norm = paste0(normalizeded_dir, "HTAN.rds" )




# there are 5 samples 
raw_dir = "/projects/lab.mis/elliot/JMML_project1/raw/HTAN/scRNAseq/raw/"
samples = list.dirs(raw_dir)
samples = samples[ grepl ( "filtered", samples)]

normal_df = data.frame()
for ( s in samples ){
  id = str_match(s, ".*\\/(.*)\\/filtered")[, 2]
  normal_df = rbind ( normal_df , data.frame ( pid=id, dir=s))
}




```
 
```{r}
mt_cut = 15 
nfeature_min = 200 
nfeature_max = 4000


```

# Summary {.tabset}

  * The goal of this is to prepare normal blood sample from, Chen et al, 2022
    + https://www.biorxiv.org/content/10.1101/2020.12.06.413930v1
  * There are 5 samples that we downloaded from the HTAN website
  * We run Prepocessing and QC to filter out low quality cells. 
  * alls samples were merged for downstream analysis 

## pre-processing workflow {.tabset}

  * The steps below encompass the standard pre-processing workflow based on Seurat's Basic recommendations. However some paremeters might be tweaked for the specific experiments. 
  
  * Cells are filtered based on 2 main factors.
  
    + cells that have unique feature counts over `r nfeature_max` or less than `r nfeature_min`
    
      - Low-quality cells or empty droplets will often have very few genes
      -  Cell doublets or multiplets may exhibit an aberrantly high gene count
      - __Thus we try to find a Goldilock range: not too low but not too high__
    
    + cells with > `r mt_cut `% mitochondrial counts. High mitochondrial indicates Low-quality / dying cells.
    
      -Low-quality / dying cells often exhibit extensive mitochondrial contamination
      - We calculate mitochondrial QC metrics with the `PercentageFeatureSet` function, which calculates the percentage of counts originating from a set of features
      - We use the set of all genes starting with `MT-` as a set of mitochondrial genes

```{r}

# preprocess each of the samples 
# each sample will be store din a list 

slist = list () # each sample will be store in this list to be merged later. 
preplots = list()
normal_df$before_preprocess = 1
normal_df$after_preprocess = 1



for ( r in 1:nrow ( normal_df) ){

# r=1 
pid = normal_df[r:r, ]
id = pid$pid
# Load the data
data <- Read10X(data.dir = pid$dir )
# Initialize the Seurat object with the raw (non-normalized data).
pbmc <- CreateSeuratObject(counts = data, project = pid$pid, min.cells = 3, min.features = 200)
# add mitochondria percent 
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
head(pbmc@meta.data, 5)

# plot mitochondria 
mt_plot <- plot_hist_perc ( pbmc@meta.data, plot_this="percent.mt", ccc="#6e9be5" , cut_here = 15, msg = "" , binwidth=5, y=.3) 
nfeature_plot <- plot_hist_perc ( pbmc@meta.data, plot_this="nFeature_RNA", ccc="#6e9be5" 
                                  , cut_here = c (nfeature_min,  nfeature_max)
                                  , msg = "" 
                                  , binwidth=200, y=.02  # y is the position of annotated text
                                  ) 

preplots[[id]][["hist"]] <- (  ( mt_plot + ggtitle(pid$pid) ) / nfeature_plot )

# quantile ( pbmc@meta.data$percent.mt) # here we see that 50% of the cells will be removed if MT is set to 10% 
# quantile ( pbmc@meta.data$nFeature_RNA) # 25% will be removed if the high end is set to 2500 



# Visualize QC metrics as a violin plot
before_plot = VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, cols="#FAF632" )
before_total = length ( Cells(pbmc) )

# after 
pbmc =  subset(pbmc, subset = nFeature_RNA > nfeature_min & nFeature_RNA < nfeature_max & percent.mt < mt_cut )
after_total = length ( Cells(pbmc) )
after_plot = VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, cols="#526E9F")


preplots[[id]][["befor_after"]]  =  (before_plot  ) / after_plot 

normal_df[r, ]$before_preprocess = before_total
normal_df[r, ]$after_preprocess = after_total

# save seurat object
slist[[id]] = pbmc

print ( paste ( "completed", id ))  
}


```

### tabulate. 

  * before and after shows how many cells were kept after preprocessing. 


```{r fig=TRUE,fig.width=16, fig.height=10, echo=FALSE, include=TRUE, results='asis' }
temp = normal_df 
temp$dir = NULL 

kable( temp, format = "html" , row.names = F, caption = "Samples" ) %>% kable_classic(full_width = F, position = "center")

```


```{r fig=TRUE,fig.width=17, fig.height=10, echo=FALSE, include=TRUE, results='asis' }

for ( n in names ( preplots)){
  cat ( "### ", n, "\n\n"  )
  print ( ( preplots[[n]]$hist)  |  (preplots[[n]]$befor_after ))
  cat ( '\n\n')
}

```


```{r}
# final merging 

names ( slist )




pbmc.big <-  merge(slist[["HTA4_26" ]]
                  , y = c(slist[["HTA4_27"]], slist[["HTA4_28"]]
                          , slist[[ "HTA4_29" ]], slist[[ "HTA4_30" ]])
                  
                  , add.cell.ids = c("HTA4_26", "HTA4_27", "HTA4_28"
                                     , "HTA4_29" , "HTA4_30"
                                     ), project = "Normal_chen_2022")


saveRDS(pbmc.big, filename  ) # we save this before normalizing because we may need normalize again when combining with JMML samples 

unique ( pbmc.big@meta.data$orig.ident)

# this are the data 
head  ( pbmc.big@assays$RNA@data[ 1:3, 1:3] )
head ( pbmc.big@meta.data)
cell =  pbmc.big@meta.data$orig.ident


pbmc.big[["exp"]] <- cell
# here we need to also correct for batch and mitochondria. Cell cycles 
pbmc.big <- SCTransform(pbmc.big, vst.flavor = "v2", verbose = TRUE, vars.to.regress=  c("percent.mt","exp")  ) 
# save before clustering... does'nt matter but it shoul save some space. 
saveRDS( pbmc.big, filename_norm ) 

 pbmc.big <-    pbmc.big %>% 
    # ScaleData(features = rownames(pbmc)) %>%, not needed since SCTransform should had done this already 
    # Scaled data is here pbmc.big[["SCT"]]@scale.data[1:3, 1:3 ]
    RunPCA(npcs = 30, verbose = FALSE) %>%
    RunUMAP(reduction = "pca", dims = 1:30, verbose = FALSE) %>%
    FindNeighbors(reduction = "pca", dims = 1:30, verbose = FALSE) %>%
    FindClusters(resolution = 0.7, verbose = FALSE)
# this is how many clusters this thing yield 
clusters = levels ( pbmc.big )
ccc_clusters = getPalette( length ( clusters) )


DimPlot(pbmc.big, label = T, repel = T ) # plots total - default for seurat transformation is 1:30 so no need to mess with this. 
# changing resolution does'nt seem to effect this experiment

# want to visualize if there is a batch 

batch_plot = DimPlot(pbmc.big, label = T, repel = T, group.by = "orig.ident") + 
  ggtitle("Unsupervised clustering") +  scale_color_manual(values = brewer.pal(8, "Dark2") )

```

```{r}

# auto assign 

source("../ScType/gene_sets_prepare.R")
source("../ScType/sctype_score_.R")
db_ = "../ScType/ScTypeDB_full.xlsx";
tissue = "Immune system" # e.g. Immune system, Liver, Pancreas, Kidney, Eye, Brain

# prepare gene sets
gs_list = gene_sets_prepare(db_, tissue)

es.max = sctype_score(scRNAseqData = pbmc.big[["SCT"]]@scale.data, scaled = TRUE, 
                      gs = gs_list$gs_positive, gs2 = gs_list$gs_negative) 

# merge by cluster

cL_resutls = do.call("rbind", lapply(unique(pbmc.big@meta.data$seurat_clusters), function(cl){
  es.max.cl = sort(rowSums(es.max[ ,rownames(pbmc.big@meta.data[pbmc.big@meta.data$seurat_clusters==cl, ])]), decreasing = !0)
  head(data.frame(cluster = cl, type = names(es.max.cl), scores = es.max.cl, ncells = sum(pbmc.big@meta.data$seurat_clusters==cl)), 10)
}))
sctype_scores = cL_resutls %>% group_by(cluster) %>% top_n(n = 1, wt = scores)  


# set low-confident (low ScType score) clusters to "unknown"
sctype_scores$type[as.numeric(as.character(sctype_scores$scores)) < sctype_scores$ncells/4] = "Unknown"
print(sctype_scores[,1:3])


pbmc.big@meta.data$customclassif = ""
for(j in unique(sctype_scores$cluster)){
  cl_type = sctype_scores[sctype_scores$cluster==j,]; 
  pbmc.big@meta.data$customclassif[pbmc.big@meta.data$seurat_clusters == j] = as.character(cl_type$type[1])
}

ccc = setNames( getPalette( length ( unique ( unique ( pbmc.big@meta.data$customclassif)) ) )
                               , unique ( pbmc.big@meta.data$customclassif)
                               )


cell_type_plot = DimPlot(pbmc.big, reduction = "umap", label = TRUE, repel = TRUE, group.by = 'customclassif')    +  
  scale_color_manual(values = ccc )



```

```{r}

# detail of cell id 


# prepare edges
cL_resutls=cL_resutls[order(cL_resutls$cluster),]; edges = cL_resutls; edges$type = paste0(edges$type,"_",edges$cluster); edges$cluster = paste0("cluster ", edges$cluster); edges = edges[,c("cluster", "type")]; colnames(edges) = c("from", "to"); rownames(edges) <- NULL

# prepare nodes
nodes_lvl1 = sctype_scores[,c("cluster", "ncells")]; nodes_lvl1$cluster = paste0("cluster ", nodes_lvl1$cluster); nodes_lvl1$Colour = "#f1f1ef"; 
nodes_lvl1$ord = 1; nodes_lvl1$realname = nodes_lvl1$cluster; nodes_lvl1 = as.data.frame(nodes_lvl1); nodes_lvl2 = c(); 
ccolss= ccc_clusters
  
for (i in 1:length(unique(cL_resutls$cluster))){
  dt_tmp = cL_resutls[cL_resutls$cluster == unique(cL_resutls$cluster)[i], ]; nodes_lvl2 = rbind(nodes_lvl2, data.frame(cluster = paste0(dt_tmp$type,"_",dt_tmp$cluster), ncells = dt_tmp$scores, Colour = ccolss[i], ord = 2, realname = dt_tmp$type))
}

nodes = rbind(nodes_lvl1, nodes_lvl2); nodes$ncells[nodes$ncells<1] = 1;
files_db = openxlsx::read.xlsx(db_)[,c("cellName","shortName")]; files_db = unique(files_db); nodes = merge(nodes, files_db, all.x = T, all.y = F, by.x = "realname", by.y = "cellName", sort = F)
nodes$shortName[is.na(nodes$shortName)] = nodes$realname[is.na(nodes$shortName)]; nodes = nodes[,c("cluster", "ncells", "Colour", "ord", "shortName", "realname")]

mygraph <- graph_from_data_frame(edges, vertices=nodes)

# Make the graph
gggr<- ggraph(mygraph, layout = 'circlepack', weight=I(ncells)) + 
  geom_node_circle(aes(filter=ord==1,fill=I("#F5F5F5"), colour=I("#D3D3D3")), alpha=0.9) + geom_node_circle(aes(filter=ord==2,fill=I(Colour), colour=I("#D3D3D3")), alpha=0.9) +
  theme_void() + geom_node_text(aes(filter=ord==2, label=shortName, colour=I("#ffffff"), fill="white", repel = !1, parse = T, size = I(log(ncells,25)*1.5)))+ geom_node_label(aes(filter=ord==1,  label=shortName, colour=I("#000000"), size = I(3), fill="white", parse = T), repel = !0, segment.linetype="dotted")

DimPlot(pbmc.big, reduction = "umap", label = TRUE, repel = TRUE, cols = ccolss) +  gggr 



```


## Visuals {.tabset}

* merged the samples into a single assay and normalized/scaled and removed unwanted variation with the `SCTransform` function. 
* We corrected for Mitochondria content and sample. 


### Overall After scaling and normalization 

  * sample names are colored to indicate any batch effect even after corrected

```{r fig=TRUE,fig.width=10, fig.height=8, echo=FALSE, include=TRUE, results='asis' }
batch_plot
```

### Cell type

  * we used an automated cell-type indentifier ScType to label and group each clusters. 
  * Ianevski et al, 2022
  * https://www.nature.com/articles/s41467-022-28803-w


```{r fig=TRUE,fig.width=10, fig.height=8, echo=FALSE, include=TRUE, results='asis' }
cell_type_plot + ggtitle ( "cell type")
```

### Bubble 

* shows what other cell types were considered with respect to each of the clusters 

```{r fig=TRUE,fig.width=18, fig.height=10, echo=FALSE, include=TRUE, results='asis' }
DimPlot(pbmc.big, reduction = "umap", label = TRUE, repel = TRUE, cols = ccolss) +  gggr 
```


### table output. 


```{r fig=TRUE,fig.width=12, fig.height=10, echo=FALSE, include=TRUE, results='asis' }

DT::datatable(cL_resutls,
          caption = 'Cell type scores', rownames = F, 
          class = 'cell-border stripe', 
          filter = 'top',
          extensions = 'Buttons',
          fillContainer = FALSE, 
          options = list(pageLength = 20, 
                         autoWidth = F,
                         dom = 'Bfrtip',
                         buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                         scrollX = TRUE, 
            selection="multiple"
          ))


```

